Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-10-15 paper/overview.tex
Edits.

\paragraph{A Note on region identifiers} In \name, every \C{letregion}
expression introduces a new stack region with a static identifier that
uniquely identifies the region within its scope. At runtime, a
\C{letregion} expression could be evaluated multiple times if, for
instance, it is wrapped inside a \C{while} loop or a recursive method,
introducing multiple stack regions with the same static identifier.
Which among these regions the identifier is bound to under a given
runtime context is then determined by the scope of the identifier.
Furthermore, any proposition involving static region identifiers is
true if and only if the proposition is true in \emph{all} possible
runtime contexts containing bindings for those identifiers. For
instance, consider the following example:
%\begin{minipage}{1.55in}
\begin{codejava}
for (int i=0; i<=10; i++) {
  letregion R0 {
    ...
    letregion R1 {
      A a1 = new@R1 A();
    }
  }
}
\end{codejava}
%\end{minipage}
%\begin{minipage}{1.4in}
%\begin{codejava}
%void m(int i) {
%  letregion R0 {
%    A a = new@R0 A();
%  }
%  this.m(++i);
%}
%...
%this.m(1);
%\end{codejava}
%\end{minipage}
When \C{i=2}, the \C{new} expression refers to the \C{R1} (and the
corresponding stack region) introduced in the 2nd iteration of the
loop, as it is the only \C{R1} in scope. Furthermore, the proposition
\emph{\C{a1} refers to an object of type \C{A} contained in region
\C{R1}} (denoted succinctly as $\C{a1}:\C{A@R1}$) is true because it
is true for all stack regions identified by \C{R1} at runtime.
Likewise, the propostion $\C{R0} \outlives \C{R1}$ is true because it
is true in every runtime context that binds \C{R0} and \C{R1} to
regions .

Note that it is possible for multiple stack regions introduced by
evaluating a \C{letregion} multiple times in a recursive definition to
be in outlives relationship. For instance, in the following example:
\begin{codejava}
void m(A a) {
  letregion R0 {
    B b = new@R0 B();
    b.x = a;
    this.m(new@R0 A());
  }
}
\end{codejava}
stack regions identified by \C{R0} introduced in successive recursive
calls are in outlives relationship. Consequently, the assignment to
\C{b.x} should be considered safe. Region type system has to allow for
this possibility.

\subsection{Allocation Context}
\label{sec:alloc-ctxt}

As evident from the examples shown above, the suffix \C{@R} in
\C{new@R} instruct the \C{new} keyword to allocate the newly created
object in the region labeled \C{R}. However, \name does not require
the allocation region to be specified following a \C{new}. In the
absence of explicit specification, \name allocates the newly created
object in its \emph{allocation context}, which is the region
introduced by the closest enclosing \C{letregion} block. Besides being
a syntactic convenience, the facility to elide allocation region
specification is imperative to interface region-oblivious standard
library code with region-aware application code. For instance, the
following \name code, which makes explicit use of regions, calls
the standard library's iterator methods, which are region-oblivious:
\begin{center}
\begin{codejava}
  int findMin(List<int> l) {
    assert(!l.isEmpty());
    int min = l.get(0);
    letregion R0 {
      ListIterator<int> iter = l.listIterator();
      while(iter.hasNext()) {
        int n = iter.getNext();
        if (n<min) min = n;
      }
    }
    return min;
  }
\end{codejava}
\end{center}
Due to the absence of new region creations, and explicit region
annotations with \C{new} expressions, all new objects created by a
standard library method are allocated in the allocation context for
the method call. In the above example, the allocation context for the
call to \C{List} library's \C{listIterator} method is the region
\C{R1}. Therefore, all objects created by the method, including the
iterator object, are allocated in \C{R0}. This fact should be captured
appropriately by the region type of \C{listIterator} method if the
type system were to prove the safety of the above code.

@@

In \name, like in \csharp, values of some primitive types, such as
integers and bools, are unboxed, whereas others, such as strings are
arrays, are boxed. Unlike unboxed values, boxed values are objects,
hence passed by reference. It is therefore important to be mindful
of allocation regions for strings and arrays, but not so much for
integers and bools. For instance, in the following code, the
assignment to integer variable \C{x} is safe, but the assignment to
the string variable \C{y} is unsafe, as the allocation region (\C{R0})
of the string constant has lesser lifetime than \C{y}.
\begin{center}
\begin{codejava}
  int x;
  string y;
  letregion R0 {
    x = 2;
    y = "Hello, World!";
  }
  ...
\end{codejava}
\end{center}

@@

\name's stack regions have lexically scoped lifetimes, hence are not
suitable for applications that need to transfer data between
autonomous actors, such as \naiad dataflow operators. To support these
applications, it is imperative to support regions whose lifetimes
transcend the boundaries of lexical blocks, functions, or even
programs. Regions that fit this description in \name are called
\emph{dynamic regions}, or \emph{transferable regions}.

Transferable regions, like stack regions, are memory regions
Transferable regions are first class values of \name; they are
considered objects of \C{Region} class, and, like objects of other
classes, they are created using the \C{new} keyword, can be passed as
arguments, and returned from methods. Like a stack region, a
transferable region facilitates arbitrary object allocations, but
unlike a stack region, it has a well-defined \emph{root} object
denoting the root of the data structure being transferred. In a
typical dataflow computation, an upstream actor (e.g., a \C{SELECT}
operator) constructs a transferable region, sets its root to the data
structure supposed to contain intermediate results, populates the data
structure with intermediate results, and then transfers it to a
downstream actor (e.g., a \C{COUNT} operator), which performs further
processing. Since a transferable region escapes the lifetime of the
sender, there must be no references from inside of the transferable
region to objects allocated in other memory regions of the sender.
Such references, if exist, may become invalid references in the
context of recipient's address space, jeopardizing memory safety.
\name relies on its region type system to prevent such unsafe
references from being created.

>>>>

\paragraph{A Note on region identifiers} Every stack region in \name
is associated with a static identifier that uniquely identifies the
region within its scope. At runtime, if a \C{letregion} expression is
evaluated multiple times in a loop or a recursive method, the
corresponding identifier is bound to a new stack region each time. 
Any proposition involving static region identifiers is
considered true at a program location if and only if the proposition
is true under all possible evaluation contexts of that program
location. For instance, consider the following example:
%\begin{minipage}{1.55in}
\begin{numcodejava}
for (int i=0; i<=10; i++) {
  letregion R0 {
    letregion R1 {
      A a1 = new@R1 A();
      ...
    }
  }
}
\end{numcodejava}
%\end{minipage}
%\begin{minipage}{1.4in}
%\begin{codejava}
%void m(int i) {
%  letregion R0 {
%    A a = new@R0 A();
%  }
%  this.m(++i);
%}
%...
%this.m(1);
%\end{codejava}
%\end{minipage}
The identifiers \C{R0} and \C{R1} are bound to new stack regions each
time the loop is evaulated. Nonetheless, the propositions (a). $\C{R0}
\outlives \C{R1}$, and (b). $\C{a1}:\C{A@R1}$ (read as \emph{\C{a1}
refers to an object of type \C{A} contained in region \C{R1}}) are
true at line 5, as they are true under all possible bindings of \C{R0}
and \C{R1} at line 5.

% Interestingly, it is possible for multiple stack regions introduced by
% evaluating a the same \C{letregion} expression to be in outlives
% relationship. For instance, in the following example:
% \begin{codejava}
% void m(A a) {
%   letregion R0 {
%     B b = new@R0 B();
%     b.x = a;
%     this.m(new@R0 A());
%   }
% }
% \end{codejava}
% Stack regions identified by \C{R0} introduced in successive recursive
% calls are in outlives relationship. Consequently, the assignment to
% \C{b.x} should be considered safe. Region type system has to allow for
% this possibility.

\subsection{Allocation Context and Qualified Region Polymorphism}
\label{sec:alloc-ctxt}

As evident from the examples shown above, the suffix \C{@R} in
\C{new@R} instructs the \C{new} keyword to allocate the newly created
object in the region labeled \C{R}. However, \name does not require
the allocation region to be specified following a \C{new}. In the
absence of explicit specification, \name allocates the newly created
object in its \emph{allocation context}, which is the region at the
top of the stack when \C{new} is executed. For instance, the
\C{findMin} function shown below, which makes explicit use of regions,
calls the standard \C{List} library's \C{listIterator} method:
\begin{center}
\begin{codejava}
  int findMin(List<int> l) {
    assert(!l.isEmpty());
    int min = l.get(0);
    letregion R0 {
      ListIterator<int> iter = l.listIterator();
      while(iter.hasNext()) {
        int n = iter.getNext();
        if (n<min) min = n;
      }
    }
    return min;
  }
\end{codejava}
\end{center}
The \C{listIterator} method neither creates new regions, nor does it
annotate \C{new} expressions with region specifications. Consequently,
all new objects created by the method are allocated in the allocation
context for the call to \C{listIterator}, which in this case is
\C{R0}.

Observe that \C{listIterator} can be called under multiple different
allocation contexts, and each time it returns a \C{ListIterator}
allocated in its allocation context. The iterator object might hold
references to the list, requiring the list to be allocated in a region
that outlives \C{listIterator}\!'s allocation context. However, modulo
this constraint, \C{listIterator} is not concerned about where the
list is allocated. As such, \C{listIterator} is
\emph{region-polymorphic} with respect to (a). its allocation context
argument, and (b). the allocation region of the list, subject to the
constraint that the later outlives the former. We call such region
polymorphism with constraints in \name as \emph{qualified region
polymorphism}. The provision to elide allocation region
specifications, and the ability to infer qualified region-region
polymorphic types are pivotal to interface region-oblivious standard
library code with region-aware application code in \name.

@@

In \name, like in \csharp, values of some primitive types, such as
integers and bools, are unboxed, whereas others, such as strings are
arrays, are boxed. Unlike unboxed values, boxed values are objects,
hence passed by reference. It is therefore important to be mindful
of allocation regions for strings and arrays, but not so much for
integers and bools. For instance, in the following code, the
assignment to integer variable \C{x} is safe, but the assignment to
the string variable \C{y} is unsafe, as the allocation region (\C{R0})
of the string constant has lesser lifetime than \C{y}.
\begin{center}
\begin{codejava}
  int x;
  string y;
  letregion R0 {
    x = 2;
    y = "Hello, World!";
  }
  ...
\end{codejava}
\end{center}
Accordingly, \name's region type system lets the former assignment
pass by, but flags the later as an error.

@@

\name's stack regions have lexically scoped lifetimes, hence are not
suitable for applications that need to transfer data between
autonomous entities, such as \naiad dataflow operators. To support
these applications, it is imperative to support regions whose
lifetimes transcend the boundaries of lexical blocks, functions, or
even programs. Regions that fit this description in \name are called
\emph{dynamic regions}, or \emph{transferable regions}.

Transferable regions, like stack regions, are memory regions, but
unlike stack regions, they are first class values of \name; they are
considered objects of \C{Region} class, and, like objects of other
classes, they are created using the \C{new} keyword, can be passed as
arguments, stored in data structures, and returned from methods. Like
a stack region, a transferable region facilitates arbitrary object
allocations, but unlike a stack region, it has a well-defined
\emph{root} object denoting the root of the data structure being
transferred. In a typical dataflow computation, an upstream actor
(e.g., a \C{SELECT} operator) constructs a transferable region, sets
its root to the data structure containing intermediate results, and
then transfers it to a downstream actor (e.g., a \C{COUNT} operator),
which performs further processing. Since a transferable region escapes
the lifetime of the sender, there must be no references from inside of
the transferable region to objects allocated in other memory regions
of the sender. Such references, if exist, may become invalid
references in the context of recipient's address space, jeopardizing
memory safety. \name relies on its region type system to prevent such
unsafe references from being created.

### 2015-10-15 paper/type-system.tex
Edits.

\FB class definitions can be understood as extensions of FGJ class
definitions with region annotations and region parameters. More
precisely, a \fbname definition of a class is obtained from its
Featherweight Generic Java definition by (a). lifting all core types
in the body of the class (i.e, fields, constructors and methods) to
their region-annotated versions, and (b). adding region parameters to
its header. Note that region parameterization is independent of its
parameterization over core types; they are not conflated to make the
class definition parameteric over region types. We let $a, b$ range
over (core) type parameters and $\rho$ range over region parameters of
a class. The following definition of \C{Pair} class is an example of
type and region parametric definition in \FB (The symbol $\extends$
should be read \emph{extends}):
\begin{codejava}[mathescape=true]
class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
                      \conj \rho_2 \outlives \rho^a$>
          <a $\extends$ Object, b $\extends$ Object> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}
Every class definition in \FB is necessarily parameterized over the
allocation region of its objects. As a rule, allocation region
parameter (denoted $\rho^a$) is the first region parameter of a class.
Besides the allocation region of its objects, \C{Pair} class is also
parametric over the regions its first and second elements are
allocated in. References between objects allocated in different
regions are only allowed if the referred object is guaranteed to
outlive the referring object. In case of \C{Pair} class, this means
that allocation regions ($\rho_1$ and $\rho_2$) of both objects that
make up the pair must outlive the allocation region ($\rho^a$) of the
\C{Pair} object. Such conditions over region parameters of a class
need to be recorded in its header as region constraints ($\phi$) in
order for the class to be judged well-formed by the type system
(Fig.~\ref{fig:fb-morewfrules}).

@@

>>>>

\FB class definitions can be understood as extensions of FGJ class
definitions with region annotations and region parameters. More
precisely, a \fbname definition of a class is obtained from its
Featherweight Generic Java definition by (a). lifting all core types
in the body of the class (i.e, fields, constructors and methods) to
their region-annotated versions, and (b). adding region parameters to
its header. Note that region parameterization is independent of its
parameterization over core types; they are not conflated to make the
class definition parametric over region types. We let $a, b$ range
over (core) type parameters and $\rho$ range over region parameters of
a class. The following definition of \C{Pair} class is an example of
type and region parametric definition in \FB (The symbol $\extends$
should be read \emph{extends}):
\begin{codejava}[mathescape=true]
class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
                      \conj \rho_2 \outlives \rho^a$>
          <a $\extends$ Object, b $\extends$ Object> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}
Every class definition in \FB is necessarily parameterized over the
allocation region of its objects. As a rule, allocation region
parameter (denoted $\rho^a$) is the first region parameter of a class.
Besides the allocation region of its objects, \C{Pair} class is also
parametric over the regions its first and second elements are
allocated in. References between objects allocated in different
regions are only allowed if the referred object is guaranteed to
outlive the referring object. In case of \C{Pair} class, this means
that allocation regions ($\rho_1$ and $\rho_2$) of both objects that
make up the pair must outlive the allocation region ($\rho^a$) of the
\C{Pair} object. Such conditions over region parameters of a class
need to be recorded in its header as region constraints ($\phi$) in
order for the class to be judged well-formed by the type system
(Fig.~\ref{fig:fb-morewfrules}).

@@

The expression language admits expressions to instantiate
($e\inang{\ralloc\rbar}$) region parameters, and to apply
($e(\bar{e})$) functions. For a method call, however, region parameter
instantiation and method application are captured in a single
expression ($e.mn\inang{\ralloc \rbar}(\ebar)$). The language also
admits a big-lambda expression, which, when used in conjunction with
the instantiation expression, serves the bureaucratic purpose of
renaming bound region variables in the type of a function closure.

The
